Elaboration with First-Class Implicit Function Types

### 2020-05-10 paper.tex
revs

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular function arguments should be inferred by default. In
Agda and Coq, one can use bracketed function domains for this purpose:
\begin{alignat*}{3}
  & id : \{A : \kw{Set}\}\to A \to A \hspace{5em} \kw{Definition}\,id\,\{A : \kw{Type}\}(x : A) := x.\\
  & id\,x = x
\end{alignat*}
In GHC, one can use $\kw{forall}$ to define implicit function
types\footnote{This notation requires language extensions
  $\mathsf{KindSignatures}$ and $\mathsf{RankNTypes}$; one could also write the type
  $a \to a$ and GHC would silently insert the quantification.}
\begin{alignat*}{3}
  & id :: \kw{forall}\,(a :: \kw{*}).\,a \to a\\
  & id\,x = x
\end{alignat*}
In all of the above cases, if we apply $id$ to an argument, the implicit type
argument is provided by elaboration. For example, in Agda, $id\,true$ is
elaborated to $id\,\{Bool\}\,true$, and analogously in GHC and Coq. In all three
systems, there is also a way to explicitly specify implicit arguments: in Agda
we may put arguments in brackets as we have seen, in Coq we can prefix a name
with $\kw{@}$ to make every implicit argument explicit, as in
$\kw{@}id\,bool\,true$, and in GHC we can enable the language extension
$\mathsf{TypeApplications}$ \cite{eisenberg2016visible} and write
$id\,\kw{@}Bool\,True$.

Implicit function types are \textbf{first-class} if they can be manipulated like any
other type. Coq is an example for a system where this is \emph{not} the case.
In Coq, the core language does not have an actual implicit function type,
instead, implicitness is tied to particular \emph{names}, and while we can write
$list\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type with
polymorphic elements, the brackets here are simply ignored by Coq. For example,
Coq accepts the following definition:
\begin{alignat*}{3}
  & \kw{Definition}\,poly\,: \kw{forall}\,(f : \kw{forall}\,\{A : \kw{Type}\},\,A\to A),\,bool\,*\,nat :=\\
  & \quad \kw{fun}\,f \Rightarrow (f\,bool\,true,\,f\,nat\,0)
\end{alignat*}
This is a higher-rank polymorphic function which returns a pair. Note that $f$
is applied to two arguments, because the implicitness in $\kw{forall}\,\{A :
\kw{Type}\},\,A\to A$ is silently dropped.

In GHC Haskell, $\kw{forall}$ types are more flexible. We can write the
following, with $\kw{RankNTypes}$ enabled:
\begin{alignat*}{3}
  & poly :: (\kw{forall}\,a.\,a \to a) \to (Bool,\,Int) \\
  & poly\,f = (f\,True,\,f\,0)
\end{alignat*}
However, polymorphic types are only supported in function domains and as fields
of algebraic data constructors. We cannot instantiate an arbitrary type
parameter to a $\kw{forall}$, as in $[\kw{forall}\,a.\,a\to a]$ for a list type
with polymorphic elements. While this type is technically allowed by the
$\kw{ImpredicativeTypes}$ language extension, as of GHC 8.8 this extension is
deprecated and is not particularly usable in practice.

In Agda, implicit functions are truly a first-class notion, and we may have
$List\,(\{A : \kw{Set}\}\to A \to A)$ without issues. However, Agda's elaboration
still has limitations when it comes to handling implicit functions. Assume that
we have $[]$ for the empty list and $\blank::\blank$ for list extension,
and consider the following code:
\begin{alignat*}{3}
  & polyList : List\,(\{A : \kw{Set}\}\to A \to A)\\
  & polyList = (\lambda\,x \to x)\,::\,[]
\end{alignat*}
Agda 2.6.0.1 does not accept this. However, it does accept $polyList =
(\lambda\,\{A\}\,x \to x)\,::\,[]$. The issue is the following. Agda first
infers a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
inferred type with the given $List\,(\{A : \kw{Set}\}\to A \to A)$
annotation. However, when Agda elaborates $\lambda\,x \to x$, it does not yet
know anything about the element type of the list; it is an undetermined
unification variable. Hence, Agda does not know whether it should insert an
extra $\lambda\,\{A\}$ or not. If the element type is later found to be an
implicit function, then it should, otherwise it should not. To solve this
conundrum, Agda simply assumes that any unknown type is \emph{not} an implicit
function type, and elects to not insert a lambda. This assumption is often
correct, but sometimes --- as in the current case --- it is not.

@@

\begin{itemize}
  \item
    A metasubstitution $\theta : \Theta_0 \To \Theta_1$ assigns to each variable
    in $\Theta_1$ a term in $\Theta_0$ , hence it is represented as a list of
    terms $(\alpha_1 \mapsto t_1,\,...\,\,\alpha_i \mapsto t_i)$.
  \item
    We define the action of a metasubstitution on contexts and terms by
    recursion; we write action on contexts as $\Gamma[\theta]$ and action on
    terms as $t[\theta]$. We remark that there is no abstraction for
    metavariables in the core syntax, so we do not have to handle variable
    capture (or index shifting).
  \item
    We list some of the admissible rules in Figure \ref{fig:metasub}, which
    are relevant in the rest of the paper.
\end{itemize}

@@

\begin{example}
We return to the $\mi{polyList}$ example from Section \ref{sec:issues}. We trace
elaboration using the extended algorithm.
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}
      \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil} \\
  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)} \\
  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}\\
  & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = cons\,
                       \{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,
                       (\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\\
  & \scriptstyle{8 }  && \qquad \echeckt{nil}{(List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))}\\
  & \scriptstyle{9 }  && \qquad = nil\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\\
                       & \scriptstyle{10} && \quad = cons\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,(\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\,\\
                       & && \hspace{4.4em} (nil\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\})\\
  & && \hspace{3em} : List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{11} && \quad \unify\,(List\,(\{A : \U\}\to A \to A))\,(List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{13} && \qquad\quad \unify\,\alpha_1\,((A : \U) \TCons (\alpha_3\,A))\\
  & \scriptstyle{14} && \qquad \quad \unify\,(A \to A)\,(\{z : \ol{\alpha_3\,A}\} \to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}\to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\})\\
  & \scriptstyle{15} && \qquad \qquad \unify\,(\alpha_3\,A)\,\epsilon\\
  & \scriptstyle{16} && \qquad \qquad \unify\,(A \to A)\,(\alpha_2\,\{A\} \to \alpha_2\,\{A\})\\
  & \scriptstyle{17} && \qquad \qquad \quad \unify\,A\,(\alpha_2\,\{A\})\\
  & \scriptstyle{18} && \qquad \qquad \quad \unify\,A\, A\\
  & \scriptstyle{19} && = cons\,\{\{A : \U\}\to A \to A\}\,(\lambda\,\{A\}\,x.\,x)\,(nil\,\{\{A : \U\}\to A \to A\})
\end{alignat*}
We diverge from the previous attempt at line $\scriptstyle{5}$. Here, we check
$\lambda\,x.\,x$ with the meta $\alpha_0$, so we create a fresh $\alpha_1 :
\Tel$ meta and wrap the result as $\lambda\,\{y :
\ol{\alpha_1}\}.\,\lambda\,x.\,x$. This result has the inferred type $\{y :
\ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y :
\ol{\alpha_1}\}$, and we promptly unify $\alpha_0$ (which stands for the list
element type) with it. On line $\scriptstyle{10}$, we return the elaborated
$cons$ expression, where the list element type is made explicit. It only remains
to unify the inferred and expected types. On line $\scriptstyle{12}$ we have the
case where a curried function type is matched with an implicit function type, so
on line $\scriptstyle{13}$ we refine $\alpha_1$ to a non-empty telescope, and
proceed unifying the domains. Note that on line $\scriptstyle{14}$ we have
$\alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}$, which results from applying the
substitution $y \mapsto (A :: z)$, and computing $\alpha_2\,\{(A :: z) : \ol{(A :
  \U) \TCons (\alpha_3\,A)}\}$ further using the \LabTirName{app-$\TCons$} rule
from Figure \ref{fig:telescopes}.

@@

\begin{example}
We illustrate now the action of constancy constraints, using the example from
Section \ref{sec:superfluous}. Again, assume $Bool : \U$ and $true : Bool$.
\begin{alignat*}{3}
& \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = true\,\sin\,x}
      \hspace{21em}\\
& \scriptstyle{1} && \quad \echeckt{\_}{\U} = \alpha_0\\
& \scriptstyle{3} && \quad \echeckt{true}{\alpha_0}\\
& \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel = \alpha_1\\
& \scriptstyle{6} && \qquad \einfert{true}\,= true : Bool\\
& \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,Bool\\
& \scriptstyle{9} && \qquad\quad \unify\,\alpha_1\,\epsilon\\
& \scriptstyle{10} && \qquad \unify\,\alpha_0\,Bool\\
& \scriptstyle{11} && \quad = true\\
& \scriptstyle{12} && \quad \einfert{x}{\true} = true : Bool\\
& \scriptstyle{14} && = (\slet\,x : Bool = true\,\sin\,x) : Bool
\end{alignat*}
First we create a fresh meta $\alpha_0$ for the hole on line $\scriptstyle{2}$,
then we check $\true$ with it. Since we are checking with a meta-headed type, we
create a fresh telescope meta $\alpha_1$, then infer $Bool$ for $true$. On line
$\scriptstyle{8}$ we create a constancy constraint and immediately try to solve
it, as described in Section \ref{sec:algorithmic}. Since $Bool$ does not depend
on the domain, we solve the constraint and thereby solve $\alpha_1$ as
$\epsilon$. Hence, we simply return $true$ on line $\scriptstyle{11}$, since
$\lambda\,\{x : \ol{\epsilon}\}.\,true$ computes to that, and elaboration
succeeds with the expected output.
\end{example}

@@

\begin{figure}
\begin{alignat*}{4}
  & IdTy   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && inc    && : Int \to Int                                        \\
  & single && : \{A : \U\} \to A \to List\,A                       && auto   && : IdTy \to IdTy                                      \\
  & id     && : IdTy                                               && auto'  && : \{B : \U\}\to IdTy \to B \to B                     \\
  & ids    && : List\,IdTy                                         && choose && : \{A : \U\} \to A \to A \to A                       \\
  & nil    && : \{A : \U\}\to List\,A                              && app    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
  & cons   && : \{A : \U\} \to A \to List\,A \to List\,A           && revapp && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
  & head   && : \{A : \U\} \to List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
  & tail   && : \{A : \U\} \to List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
  & map    && : \{A\,B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool
\end{alignat*}
\caption{Types used in Figure \ref{fig:benchmark}.}
\label{fig:benchmark_types}
\end{figure}

>>>>

\textbf{Implicit function types} are a common mechanism for conveying to the
compiler that particular function arguments should be inferred by default. In
Agda and Coq, one can use bracketed function domains for this purpose:
\begin{alignat*}{3}
  & \mi{id} : \{A : \kw{Set}\}\to A \to A \hspace{5em} \kw{Definition}\,\mi{id}\,\{A : \kw{Type}\}(x : A) := x.\\
  & \mi{id}\,x = x
\end{alignat*}
In GHC, one can use $\kw{forall}$ to define implicit function
types\footnote{This notation requires language extensions
  $\mathsf{KindSignatures}$ and $\mathsf{RankNTypes}$; one could also write the type
  $a \to a$ and GHC would silently insert the quantification.}
\begin{alignat*}{3}
  & id :: \kw{forall}\,(a :: \kw{*}).\,a \to a\\
  & id\,x = x
\end{alignat*}
In all of the above cases, if we apply $id$ to an argument, the implicit type
argument is provided by elaboration. For example, in Agda, $\mi{id}\,\mi{true}$ is
elaborated to $\mi{id}\,\{\mi{Bool}\}\,\mi{true}$, and analogously in GHC and Coq. In all three
systems, there is also a way to explicitly specify implicit arguments: in Agda
we may put arguments in brackets as we have seen, in Coq we can prefix a name
with $\kw{@}$ to make every implicit argument explicit, as in
$\kw{@}\mi{id}\,\mi{bool}\,\mi{true}$, and in GHC we can enable the language extension
$\mathsf{TypeApplications}$ \cite{eisenberg2016visible} and write
$\mi{id}\,\kw{@}\mi{Bool}\,\mi{True}$.

Implicit function types are \textbf{first-class} if they can be manipulated like any
other type. Coq is an example for a system where this is \emph{not} the case.
In Coq, the core language does not have an actual implicit function type,
instead, implicitness is tied to particular \emph{names}, and while we can write
$\mi{list}\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type with
polymorphic elements, the brackets here are simply ignored by Coq. For example,
Coq accepts the following definition:
\begin{alignat*}{3}
  & \kw{Definition}\,poly\,: \kw{forall}\,(f : \kw{forall}\,\{A : \kw{Type}\},\,A\to A),\,\mi{bool}\,*\,\mi{nat} :=\\
  & \quad \kw{fun}\,f \Rightarrow (f\,\mi{bool}\,\mi{true},\,f\,\mi{nat}\,0)
\end{alignat*}
This is a higher-rank polymorphic function which returns a pair. Note that $f$
is applied to two arguments, because the implicitness in $\kw{forall}\,\{A :
\kw{Type}\},\,A\to A$ is silently dropped.

In GHC Haskell, $\kw{forall}$ types are more flexible. We can write the
following, with $\kw{RankNTypes}$ enabled:
\begin{alignat*}{3}
  & \mi{poly} :: (\kw{forall}\,a.\,a \to a) \to (\mi{Bool},\,\mi{Int}) \\
  & \mi{poly}\,f = (f\,\mi{True},\,f\,0)
\end{alignat*}
However, polymorphic types are only supported in function domains and as fields
of algebraic data constructors. We cannot instantiate an arbitrary type
parameter to a $\kw{forall}$, as in $[\kw{forall}\,a.\,a\to a]$ for a list type
with polymorphic elements. While this type is technically allowed by the
$\kw{ImpredicativeTypes}$ language extension, as of GHC 8.8 this extension is
deprecated and is not particularly usable in practice.

In Agda, implicit functions are truly a first-class notion, and we may have
$List\,(\{A : \kw{Set}\}\to A \to A)$ without issues. However, Agda's elaboration
still has limitations when it comes to handling implicit functions. Assume that
we have $[]$ for the empty list and $\blank::\blank$ for list extension,
and consider the following code:
\begin{alignat*}{3}
  & \mi{polyList} : List\,(\{A : \kw{Set}\}\to A \to A)\\
  & \mi{polyList} = (\lambda\,x \to x)\,::\,[]
\end{alignat*}
Agda 2.6.0.1 does not accept this. However, it does accept $\mi{polyList} =
(\lambda\,\{A\}\,x \to x)\,::\,[]$. The issue is the following. Agda first
infers a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
inferred type with the given $List\,(\{A : \kw{Set}\}\to A \to A)$
annotation. However, when Agda elaborates $\lambda\,x \to x$, it does not yet
know anything about the element type of the list; it is an undetermined
unification variable. Hence, Agda does not know whether it should insert an
extra $\lambda\,\{A\}$ or not. If the element type is later found to be an
implicit function, then it should, otherwise it should not. To solve this
conundrum, Agda simply assumes that any unknown type is \emph{not} an implicit
function type, and elects to not insert a lambda. This assumption is often
correct, but sometimes --- as in the current case --- it is not.

@@

\begin{itemize}
  \item
    A metasubstitution $\theta : \Theta_0 \To \Theta_1$ assigns to each variable
    in $\Theta_1$ a term in $\Theta_0$, hence it is represented as a list of
    terms $(\alpha_1 \mapsto t_1,\,...\,\,\alpha_i \mapsto t_i)$.
  \item
    We define the action of a metasubstitution on contexts and terms by
    recursion; we write action on contexts as $\Gamma[\theta]$ and action on
    terms as $t[\theta]$. We remark that there is no abstraction for
    metavariables in the core syntax, so we do not have to handle variable
    capture (or index shifting).
  \item
    We list some of the admissible rules in Figure \ref{fig:metasub}, which
    are relevant in the rest of the paper.
\end{itemize}

@@

\begin{example}
We return to the $\mi{polyList}$ example from Section \ref{sec:issues}. We trace
elaboration using the extended algorithm.
  & \scriptstyle{0 }\qquad\qquad && \echeckt{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A : \U\}\to A \to A))}
      \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}} \\
  & \scriptstyle{2 }  && \qquad \einfert{\mi{cons}\,(\lambda\,x.\,x)} \\
  & \scriptstyle{3 }  && \qquad\quad \einfert{\mi{cons}}\\
  & \scriptstyle{4 }  && \qquad\quad = \mi{cons}\,\{\alpha_0\}\,:\,\alpha_0 \to \mi{List}\,\alpha_0 \to \mi{List}\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = \mi{cons}\,
                       \{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,
                       (\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\\
  & \scriptstyle{8 }  && \qquad \echeckt{\mi{nil}}{(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))}\\
  & \scriptstyle{9 }  && \qquad = \mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\\
                       & \scriptstyle{10} && \quad = \mi{cons}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,(\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\,\\
                       & && \hspace{4.4em} (\mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\})\\
  & && \hspace{3em} : \mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{11} && \quad \unify\,(\mi{List}\,(\{A : \U\}\to A \to A))\,(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{13} && \qquad\quad \unify\,\alpha_1\,((A : \U) \TCons (\alpha_3\,A))\\
  & \scriptstyle{14} && \qquad \quad \unify\,(A \to A)\,(\{z : \ol{\alpha_3\,A}\} \to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}\to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\})\\
  & \scriptstyle{15} && \qquad \qquad \unify\,(\alpha_3\,A)\,\epsilon\\
  & \scriptstyle{16} && \qquad \qquad \unify\,(A \to A)\,(\alpha_2\,\{A\} \to \alpha_2\,\{A\})\\
  & \scriptstyle{17} && \qquad \qquad \quad \unify\,A\,(\alpha_2\,\{A\})\\
  & \scriptstyle{18} && \qquad \qquad \quad \unify\,A\, A\\
  & \scriptstyle{19} && = \mi{cons}\,\{\{A : \U\}\to A \to A\}\,(\lambda\,\{A\}\,x.\,x)\,(\mi{nil}\,\{\{A : \U\}\to A \to A\})
We diverge from the previous attempt at line $\scriptstyle{5}$. Here, we check
$\lambda\,x.\,x$ with the meta $\alpha_0$, so we create a fresh $\alpha_1 :
\Tel$ meta and wrap the result as $\lambda\,\{y :
\ol{\alpha_1}\}.\,\lambda\,x.\,x$. This result has the inferred type $\{y :
\ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y :
\ol{\alpha_1}\}$, and we promptly unify $\alpha_0$ (which stands for the list
element type) with it. On line $\scriptstyle{10}$, we return the elaborated
$\mi{cons}$ expression, where the list element type is made explicit. It only remains
to unify the inferred and expected types. On line $\scriptstyle{12}$ we have the
case where a curried function type is matched with an implicit function type, so
on line $\scriptstyle{13}$ we refine $\alpha_1$ to a non-empty telescope, and
proceed unifying the domains. Note that on line $\scriptstyle{14}$ we have
$\alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}$, which results from applying the
substitution $y \mapsto (A :: z)$, and computing $\alpha_2\,\{(A :: z) : \ol{(A :
  \U) \TCons (\alpha_3\,A)}\}$ further using the \LabTirName{app-$\TCons$} rule
from Figure \ref{fig:telescopes}.

@@

\begin{example}
We illustrate now the action of constancy constraints, using the example from
Section \ref{sec:superfluous}. Again, assume $\mi{Bool} : \U$ and $\mi{true} : \mi{Bool}$.
\begin{alignat*}{3}
& \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = \mi{true}\,\sin\,x}
      \hspace{21em}\\
& \scriptstyle{1} && \quad \echeckt{\_}{\U} = \alpha_0\\
& \scriptstyle{3} && \quad \echeckt{\mi{true}}{\alpha_0}\\
& \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel = \alpha_1\\
& \scriptstyle{6} && \qquad \einfert{\mi{true}}\,= \mi{true} : \mi{Bool}\\
& \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,\mi{Bool}\\
& \scriptstyle{9} && \qquad\quad \unify\,\alpha_1\,\epsilon\\
& \scriptstyle{10} && \qquad \unify\,\alpha_0\,\mi{Bool}\\
& \scriptstyle{11} && \quad = \mi{true}\\
& \scriptstyle{12} && \quad \einfert{x}{\mi{true}} = \mi{true} : \mi{Bool}\\
& \scriptstyle{14} && = (\slet\,x : \mi{Bool} = \mi{true}\,\sin\,x) : \mi{Bool}
\end{alignat*}
First we create a fresh meta $\alpha_0$ for the hole on line $\scriptstyle{2}$,
then we check $\mi{true}$ with it. Since we are checking with a meta-headed type, we
create a fresh telescope meta $\alpha_1$, then infer $\mi{Bool}$ for
$\mi{true}$. On line $\scriptstyle{8}$ we create a constancy constraint and
immediately try to solve it, as described in Section
\ref{sec:algorithmic}. Since $\mi{Bool}$ does not depend on the domain, we solve
the constraint and thereby solve $\alpha_1$ as $\epsilon$. Hence, we simply
return $\mi{true}$ on line $\scriptstyle{11}$, since $\lambda\,\{x :
\ol{\epsilon}\}.\,\mi{true}$ computes to that, and elaboration succeeds with the
expected output.
\end{example}

@@

\begin{figure}
\begin{alignat*}{4}
  & \mi{IdTy}   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && \mi{inc}    && : \mi{Int} \to \mi{Int}                                        \\
  & \mi{single} && : \{A : \U\} \to A \to \mi{List}\,A                        && \mi{auto}   && : IdTy \to IdTy                                      \\
  & \mi{id}     && : IdTy                                                && \mi{auto'}  && : \{B : \U\}\to IdTy \to B \to B                     \\
  & \mi{ids}    && : \mi{List}\,IdTy                                          && \mi{choose} && : \{A : \U\} \to A \to A \to A                       \\
  & \mi{nil}    && : \{A : \U\}\to \mi{List}\,A                               && \mi{app}    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
  & \mi{cons}   && : \{A : \U\} \to A \to \mi{List}\,A \to \mi{List}\,A            && \mi{revapp} && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
  & \mi{head}   && : \{A : \U\} \to \mi{List}\,A \to A                        && \mi{runST}  && : \{A : \U\} \to (\{S : \U\} \to \mi{ST}\,S\,A) \to A     \\
  & \mi{tail}   && : \{A : \U\} \to \mi{List}\,A \to \mi{List}\,A                  && \mi{argST}  && : \{S : \U\}\to \mi{ST}\,S\,\mi{Int}                           \\
  & \mi{map}    && : \{A\,B : \U\} \to (A \to B) \to \mi{List}\,A \to \mi{List}\,B && \mi{poly}   && : IdTy \to \mi{Pair}\,\mi{Int}\,\mi{Bool}
\end{alignat*}
\caption{Types used in Figure \ref{fig:benchmark}.}
\label{fig:benchmark_types}
\end{figure}

\begin{figure}
\begin{tabular}{|lll|}
\hline
A & \multicolumn{2}{l|}{\textsc{polymorphic instantiation}}\\
\hline
A1 & $\lambda\,x\,y.\,y$ & \yesst\\
   & \text{we infer a type which may be solved to $\{A\,B : \U\}\to A \to B \to A$} &\\
   & \text{but not to $\{A : \U\}\to A \to \{B : \U\} \to B \to A$} & \\
A2 & $\mi{choose}\,\mi{id}$ & \yesst \\
A3 & $\mi{choose}\,\mi{nil}\,\mi{ids}$ & \yes \\
A4 & $\lambda\,(x : \{A : \U\}\to A \to A).\,x\,x$ & \yesst\\
A5 & $\mi{id}\,\mi{auto}$ & \yes\\
A6 & $\mi{id}\,\mi{auto'}$ & \yesst\\
A7 & $\mi{choose}\,\mi{id}\,\mi{auto}$ & \yes\\
A8 & $\mi{choose}\,\mi{id}\,\mi{auto'}$ & \no\\
A9 & $\lambda\,(f : \{A : \U\}\to(A\to A)\to \mi{List}\,A \to A).\,f\,(\mi{choose}\,\mi{id})\,\mi{ids}$ & \yesst\\
A10 & $\mi{poly}\,\mi{id}$ & \yes\\
A11 & $\mi{poly}\,(\lambda\,x.\,x)$ & \yes\\
\hline
B & \multicolumn{2}{l|}{\textsc{inference of polymorphic arguments}}\\
\hline
B1 & $\lambda\,f.\,\mi{pair}\,(f\,\mi{zero})\,(f\,\mi{true})$ & \no\\
B2 & $\lambda\,xs.\,\mi{poly}\,(\mi{head}\,xs)$ & \no\\
\hline
C & \multicolumn{2}{l|}{\textsc{functions on polymorphic lists}}\\
\hline
C1  & $\mi{length}\,\mi{ids}$ & \yes \\
C2  & $\mi{tail}\,\mi{ids}$ & \yes \\
C3  & $\mi{head}\,\mi{ids}$ & \yesst \\
C4  & $\mi{single}\,\mi{id}$ & \yesst \\
C5  & $\mi{cons}\,\mi{id}\,\mi{ids}$ & \yes \\
C6  & $\mi{cons}\,(\lambda\,x.\,x)\,\mi{ids}$ & \yes \\
C7  & $\mi{append}\,(\mi{single}\,\mi{inc})\,(\mi{single}\,\mi{id})$ & \yes \\
C8  & $\lambda\,(g:\{A:\U\}\to \mi{List}\,A \to \mi{List}\,A \to A).\,g\,(\mi{single}\,\mi{id})\,\mi{ids}$ & \yesst \\
C9  & $\mi{map}\,\mi{poly}\,(\mi{single}\,\mi{id})$ & \yes \\
C10 & $\mi{map}\,\mi{head}\,(\mi{single}\,\mi{ids})$ & \yes \\
\hline
D & \multicolumn{2}{l|}{\textsc{application functions}}\\
\hline
D1 & $\mi{app}\,\mi{poly}\,\mi{id}$        & \yes \\
D2 & $\mi{revapp}\,\mi{id}\,\mi{poly}$     & \yes \\
D3 & $\mi{runST}\,\mi{argST}$              & \yes \\
D4 & $\mi{app}\,\mi{runST}\,\mi{argST}$    & \yes \\
D5 & $\mi{revapp}\,\mi{argST}\,\mi{runST}$ & \yes \\
\hline
E & \multicolumn{2}{l|}{\textsc{$\eta$-expansion}}\\
\hline
& \multicolumn{2}{l|}{assuming $k : \{A : \U\}\to A \to \mi{List}\,A \to A$,} \\
& \multicolumn{2}{l|}{$h : \mi{Int} \to \{A:\U\}\to A \to A$}\\
& \multicolumn{2}{l|}{and $lst : \mi{List}\,(\{A:\U\} \to \mi{Int} \to A \to A)$} \\
E1 & $k\,h\,lst$ & \no \\
E2 & $k\,(\lambda\,x.\,h\,x)\,lst$ & \yesst \\
E3 & $\lambda\,(r : (\{A :\U\}\to A \to \{B : \U\}\to B \to B) \to \mi{Int}).\,r\,(\lambda\,x\,y.\,y)$ & \yes \\
\hline
\multicolumn{3}{l}{\text{{\footnotesize
``Yes*'' means that our system can infer a type for the
expression, but the lack of}}}\\
\multicolumn{3}{l}{\text{{\footnotesize
let-generalization yields unsolved metas in the type.}}}
\end{tabular}
